Tropical Geometry and the Lambda Calculus

### 2023-01-08 main.tex
commit

\author{\IEEEauthorblockN{Davide Barbarossa}
\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
Bologna, Italy \\
davide.barbarossa2@unibo.it}
\and
\IEEEauthorblockN{Paolo Pistone}
\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
Rome, Italy\\
paolo.pistone@uniroma3.it}
}

@@

\section{Tropical Geometry}
\input{section2}

\section{The Tropical Relational Model}
\input{section3}

\section{Analytic Results}
\input{section4}

\section{Applications}
\input{section5}

\section{Quasi-metric Spaces and $\Lawv$-Modules}
\input{section6}

\section{Related Work}
\input{section7}

\section{Conclusion}
\input{section8}

>>>>

\author{\IEEEauthorblockN{Davide Barbarossa}
\IEEEauthorblockA{\textit{Universit\`a di Bologna}\\
Bologna, Italy \\
davide.barbarossa@unibo.it}
\and
\IEEEauthorblockN{Paolo Pistone}
\IEEEauthorblockA{\textit{Universit\`a Roma Tre}\\
Rome, Italy\\
paolo.pistone@uniroma3.it}
}

@@

\section{Two Quantitative Approaches to the $\lambda$-calculus}
\input{section2}

\section{Tropical Linear Algebra and Linear Programs}
\input{section3}

\section{Tropical Polynomials and Non-Linear Programs  }
\input{section4}

\section{Analytic Results}
\input{section5}

\section{Applications}
\input{section6}

\section{Quasi-metric Spaces and $\Lawv$-Modules}
\input{section7}

\section{Related Work}
\input{section8}

\section{Conclusion}

\input{section9}

### 2023-01-08 section2.tex
commit

- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.

- tropical linear algebra: the category $\LREL$

- tropical polynomials and tropical roots

- tropical Laurent series and tropical roots

- tropicalization of analytic functions and their roots: 
$\trop P$

- the differential operator

>>>>

Recall the two approaches with more details on lambda-calculus and on existing challenges.

- Metric models: functions as Lipschitz maps, differential privacy.

- Problem with unrestricted duplications

- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda

- Differential lambda-calculus: differential operator, Taylor expansion (vs. 1)

- resource lambda-terms, Taylor expansion (vs. 2)

- resource lambdas are polynomials, so not linear. Why we need tropical geometry here.

### 2023-01-08 section3.tex
commit

Describe the model in a parametric way over a continuous semi-ring, as in Manzonetto et al.

Define the tropicalization $\trop$ from the $\BB R$-based model
to the $\Lawv$-based model.

Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
$$
A::= o \mid !_{n}A \multimap A
$$
The interpretation of a bounded type is defined as 
$\model A= \Lawv^{\nodel A}$, where 
$\nodel A$ is defined by $\nodel o=\B 1$ and 
$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.

Bounded types are interesting
because of the following proposition:
\begin{proposition}
For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.

\end{proposition}
\begin{proof}
It suffices to check that $\nodel A$ is finite for all bounded types $A$. Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
\end{proof}

For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
\begin{itemize}
\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;

\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.

\end{itemize}

Discuss the tropical Taylor expansion $\trop\C T(M)$.

>>>>

- matrices over a continuous semi-ring

- The tropical semi-ring $\Lawv$ coincides with the Lawvere quantale.

- tropical linear algebra: the category $\LREL$

- linear computation inside $\LREL$: linear lambda-terms through the symmetric monoidal adjunction 
$$
\LREL(Z\otimes X, Y) \simeq \LREL (Z, X\multimap Y)
$$
(with $\LREL(X,Y)\simeq \Lawv^{X\times Y}$.

### 2023-01-08 section4.tex
commit

- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric

- concavity

- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.

- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
\begin{itemize}

\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally

\item singular points: the famous example

\item tropical roots

@@

- discussion of the tropical Taylor expansion:

- characterization of the functional metric using derivatives

>>>>

- non-linear programs correspond to non-linear maps.

- tropical polynomials and tropical roots -> bounded lambda-terms
(the operation $\C M_{\leq n}(-)$ defines a graded linear exponential comonad) via the adjunction
$$
\LREL(Z\otimes !_{\leq n}X,Y) \simeq \LREL(Z, !_{\leq n}X\multimap Y) 
$$

- tropical Laurent series and tropical roots -> unbounded lambda-terms (the operation $\C M(-)$ defines a linear exponential comonad) via the adjunction
$$
\LREL(!(Z\& X),Y) \simeq \LREL(!Z, !X\multimap Y) 
$$

- tropicalization of analytic functions and their roots: 
$\trop P$

- the differential operator: formal differentiation of tropical Laurent series. Differentiation in $\LREL_{!}$.

- $\LREL_{!}$ is a differential CCC satisfying Taylor. 
Discuss the tropical Taylor expansion $\trop\C T(M)$.

Focus on \emph{bounded} simply typed $\lambda$-calculus $\BSTLC$:
$$
A::= o \mid !_{n}A \multimap A
$$
The interpretation of a bounded type is defined as 
$\model A= \Lawv^{\nodel A}$, where 
$\nodel A$ is defined by $\nodel o=\B 1$ and 
$\nodel{!_{n}A\multimap B}:= \C M_{\leq n}(\nodel A) \times \nodel B$.

Bounded types are interesting
because of the following proposition:
\begin{proposition}
For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) are polynomials.

\end{proposition}
\begin{proof}
It suffices to check that $\nodel A$ is finite for all bounded types $A$. Indeed this implies that a morphism $f:\model A\to \model B$ corresponds to a finite matrix $\widehat f: \nodel A \times \nodel B \to \Lawv$.
\end{proof}

For example (here $!_{n}(\Lawv^{X}):= \Lawv^{\C M_{\leq n}(X)}$):
\begin{itemize}
\item a map $f\in \LREL( !_{1}\Lawv, \Lawv)$ is of the form $f(x)=\min \{x+a,b\}$;

\item a map $f\in \LREL(!_{2}\Lawv, \Lawv)$ is a ``quadratic'' polynomial $f(x)=\min\{2x+a, x+b, c\}$.

@@

### 2023-01-08 section5.tex
commit

General discussion: optimization properties behave in a Lipschitz way.

- differential privacy and Lipschitzness

- log-probabilities and tropical roots

- counting computation steps (from Manzonetto, but add relational ``Lipschitz'' reasoning)

- measuring duplications of discrete functions (needs finiteness!)

>>>>

- the metric on the tensor is the usual tensor metric, the metric on the function space is the usual function metric, the metric on multisets is the multiset metric

- concavity

- Lipschitz-continuity. Highlight compositional reasoning based on Lipschitz.

- Discussion of unary tropical Laurent series, i.e.~$f:\Lawv \to \Lawv$:
\begin{itemize}

\item $f$ is locally a polynomial (i.e.~a $\min$) but not globally

\item singular points: the famous example

\item tropical roots

\end{itemize}

- discussion of the tropical Taylor expansion:

- characterization of the functional metric using derivatives
